---
title: "LRAPA PA FRM Fit"
author: "Mazama Science"
date: "2021-04-27"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MVCAA Tutorial 2: Exploring PAT Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)
```

# Background 
TO DO: discussion of PA vs FRM monitors and how lm() can be used to "correct" PA pm25 so that it more closely matches FRM pm25.

 
> PurpleAir particulate matter (PM) sensors are increasingly used in the United 
States and other countries by a variety of individuals and organizations for 
continuous monitoring of ambient air pollutant conditions, with additional 
sensors often deployed for monitoring during wildfire smoke episodes. 
The performance of these sensors must be evaluated during smoke impacted times, 
and nominally corrected for bias if necessary, to ensure accurate data are 
reported to inform appropriate health protective actions. (EPA, [PurpleAir 
PM2.5 U.S. Correction and Performance During Smoke Events 4/2020](https://cfpub.epa.gov/si/si_public_record_report.cfm?Lab=CEMM&dirEntryId=349513)) 

On the other hand, Federal Reference Method (FRM) monitors are highly reliable 
but their high cost can be limiting. 

A way to evaluate the functioning and reliability of PurpleAir (PA) sensors is 
to compare their perfomance by colocating them with FRM monitors. Moreover, 
data collected by FRM monitors could be potentially used to correct the data 
collected by PA sensors, wich tend to overestimate PM2.5 concentrations. 

By fittind a linear model having FRM data as the response, and PA data and humidity 
as the predictors (FRM_PM25 ~ PA_PM25 + humidity) we can derive always-updated 
coefficients (using datasets from the previous 3-5 days) that can then be used 
to correct PA data (FRM_PM25 fitted values = PA_PM25 corrected values).  

# LRAPA 
TO DO: discussion what LRAPA is, how they have co-located sensor-monitor pairs.
 Amazon Park -- a few plots showing that this sensor is basically good and matches the FRM data quite well.
 

> The Lane Regional Air Protection Agency (LRAPA) is a local government agency that monitors and regulates air quality in Lane County, Oregon. LRAPA is able to achieve this through its eight air monitoring stations throughout the county.
[...] LRAPA has seven air monitoring stations throughout Lane County, three of which are in Eugene. The air monitoring stations collect air samples and report the data from their respective locations. Information from these stations is used to assess levels of air pollution, identify air quality trends, and support the implementation of air pollution prevention strategies. 

For this study, we used the best performing sensor colocated with an FRM monitor (distance ~ 21 m). As you can see from the plots below, this PA sensor performed well throughout the entire 4 months period (July - October 2020) explored, and September 10-20 was the smokiest one. 

```{r Setup, eval = TRUE, warning = FALSE, message = FALSE}
# libraries 
library(MazamaCoreUtils)
library(AirSensor)
library(PWFSLSmoke)
library(lubridate)

# load customized functions (if you created a R project including the functions 
# scripts below, you can directly source them)
source("OR/LRAPA/sensorMonitorData.R")
source("OR/LRAPA/sensorMonitorFit.R")
source("OR/LRAPA/fitValueTimeSeries.R")
```

```{r Amazon-Park-assessment, eval = TRUE, warning = FALSE, message = FALSE }
# get Amazon Park sensor ID
setArchiveBaseDir(NULL)
setArchiveBaseUrl("http://data.mazamascience.com/PurpleAir/v1")
getArchiveBaseUrl() # check the url is correctly set
Amazon_Park_pas <- pas_load() %>%
  pas_filter(stringr::str_detect(label, "LRAPA-Amazon Park"))
Amazon_Park_ID <- pas_getDeviceDeploymentIDs(Amazon_Park_pas)

# create Amazon Park pat 
Amazon_Park <- pat_createNew(
  id = Amazon_Park_ID,
  label = NULL,        # not needed if you have the id
  pas = Amazon_Park_pas,  
  startdate = 20200701,
  enddate = 20201101,
  timezone = "America/Los_Angeles",
  baseUrl = "https://api.thingspeak.com/channels/",
  verbose = FALSE
  )

# evaluate Amazon Park sensor 
pat_multiplot(Amazon_Park)
pat_scatterPlotMatrix(Amazon_Park)
# pm25_A:pm25_B ~ 1 -- Great!!
# temperature:humidity ~ -0.82 (strong negative correlation) -- Great!
pat_externalFit(Amazon_Park, showPlot = TRUE)
# Adjusted R-squared:  0.9666 -- Great!!
# Smokiest period Sep 10 - 20. 
```

 
# Linear fit 
TO DO: Using lm(pm25_monitor ~ pm25 + humidity) and a 5-day window to created our model. Then use predict() to create corrected (aka "fitted") values. Code snippet to show what we are doing. For a one-week window in July, show FRM pm25, PA pm25 and corrected (PA values with the lm() calculated linear correction applied.)

```{r Linear-fit-timeseries-July, eval = TRUE, warning = FALSE, message = FALSE }
# Get monitors data 
LRAPA_monitors <- monitor_loadLatest() %>%
  monitor_subset(stateCodes = "OR") %>%
  monitor_subsetBy(countyName == "Lane")
monitorID <- Amazon_Park$meta$pwfsl_closestMonitorID

# Create dataframe (df): Amazon Park PA one-week data -- July, 2020
# NOTE: source functions in the "Setup" R block first!
sensorMonitorData_07 <-sensorMonitorData(pat = Amazon_Park,
                                        ws_monitor = LRAPA_monitors,
                                        monitorID = monitorID,
                                        startdate = 20200701,
                                        enddate = 20200706)
View(sensorMonitorData_07)

# Add pm25_monitor_fitted (aka PA pm25 corrected) to the PA df
lm <- lm(pm25_monitor ~ pm25 + humidity)
sensorMonitorData_07$pm25_monitor_fitted <- predict(lm, newdata = sensorMonitorData_07)

# Create a timeseries including PA pm25, humidity, and monitor pm25 fitted values:

# * timeseries axes
datetime <- sensorMonitorData_2$datetime
pm25_monitor <- sensorMonitorData_2$pm25_monitor # chose the one with the 
# highest values for the purpuse of plotting

# * timeseries setup
a <- approx(
  c( ymd_hms('2020-07-01T00:00:00'), ymd_hms('2020-07-06T00:00:00') ),
  n = length(datetime)
)
x.date <- as_datetime(a$y)
length(x.date)

dr <- range(x.date)
date.at <- seq(dr[1], dr[2], by="day")[seq(1,120,1)]


# * plot 
plot(datetime, pm25_monitor, xaxt="n", type="n", 
     main="Amazon Park Linear Fit: Coefficients & Fitted Values -- July 01-05, 2020",
     xlab="Date", ylab = "PM2.5" )
axis( 1, at=date.at, format(date.at,"%b %d") )
lines( x=datetime, y=sensorMonitorData_07$pm25, col = colors()[461], lwd=2)
lines( x=datetime, y=sensorMonitorData_07$pm25_monitor , col = colors()[631], lwd=2)
lines( x=datetime, y=sensorMonitorData_07$pm25_monitor_fitted, col = colors()[640], lwd=2)

## Legend:
labels <- c(
  "PA PM25 measured", "FRM PM25 measured", "FRM PM25 Fitted (PA PM25 Corrected)"
)
legend(
  "topleft",
  legend = labels,
  lty=c("solid", "solid", "solid"),
  lwd=2,
  col= c(colors()[461], colors()[631], colors()[640])
)
```


# EPA correction 
TO DO: show a 4-month plot with PA pm25, FRM pm25 and EPA "corrected" pm25 (PA values with EPA linear correction applied.); This plot should show that the EPA "over corrects" during the smoky period.
```{r}
# 4-month df including PA pm25 and FRM pm25 measured values 
sensorMonitorData_full <-sensorMonitorData(pat = Amazon_Park,
                                        ws_monitor = LRAPA_monitors,
                                        monitorID = monitorID,
                                        startdate = 20200701,
                                        enddate = 20201101)
View(sensorMonitorData_full)

# Add EPA PA pm25 corrected to sensorMonitorData_full df 
# EPA equation: PA PM2.5 corrected = 0.534*[PA_cf1(avgAB)] - 0.0844*RH +5.604
EPA_PA_PM25_corrected <- (0.534*sensorMonitorData_full$pm25 - 0.0844*sensorMonitorData_full$humidity +5.604)
sensorMonitorData_full$EPA_PA_PM25_corrected <- round(EPA_PA_PM25_corrected)

# Create 4-month timeseries including PA pm25 and FRM pm25 measured values and 
# EPA PA pm25 correceted:
# * timeseries axes
datetime <- sensorMonitorData_full$datetime
PA_pm25 <- sensorMonitorData_full$pm25 # chose the one with the 
# highest values for the purpuse of plotting

# * timeseries setup
a <- approx(
  c( ymd_hms('2020-07-01T00:00:00'), ymd_hms('2020-10-31T00:00:00') ),
  n = length(enddate)
)
x.date <- as_datetime(a$y)

dr <- range(x.date)
date.at <- seq(dr[1], dr[2], by="day")[seq(1,200,7)]

# * plot 
plot(enddate, PA_pm25, xaxt="n", type="n", 
     main="Fit Values -- July-Oct, 2020 -- 5 Days", ylim = c(-2,2),
     xlab="Date", ylab = "Values" )
axis( 1, at=date.at, format(date.at,"%b %d") )
lines( x=enddate, y= sensorMonitorData_full$pm25, col = colors()[616], lwd=2 )
lines( x=enddate, y = sensorMonitorData_full$pm25_monitor, col = colors()[131], lwd=2)
lines( x=enddate, y = sensorMonitorData_full$EPA_PA_PM25_corrected, col = colors()[636], lwd=2)

## Legend:
labels <- c(
"PA PM25 measured", "FRM PM25 measured", "EPA PA PM25 Corrected"
)
legend(
  "bottomleft",
  legend = labels,
  lty=c("solid", "solid","solid"),
  lwd=2,
  col= c(colors()[616], colors()[131], colors()[636])
)

```



# Real-time linear fit
TO DO: Show the fit parameters for EPA (straight lines) and for real-time lm(). Pick ylim() that highlights the pm25 and humidity values. Don't worry about the smoky period when the intercept goes crazy.

# Real-time vs EPA 
TO DO: Discussion and plots showing PA pm25, FRM pm25, EPA corrected pm25 and real-time corrected pm25 for one week in July and one smoky week in October
